---
title: "Scoring with KidsightsPublic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring with KidsightsPublic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

KidsightsPublic computes developmental screening scores using a Graded Response
Model (GRM) with item parameters calibrated from a large norming sample. Scores
are estimated via MAP (maximum a posteriori) optimization using CmdStan.

Two scoring functions are available:

- `score_kidsights()` — a single overall developmental score (F factor)
- `score_psychosocial()` — six psychosocial domain scores (GEN + EAT, EXT, INT,
  SLE, SOC)

## Prerequisites

CmdStan must be installed. Install it via:

```{r}
install.packages("cmdstanr", repos = c("https://stan-dev.r-universe.dev", getOption("repos")))
cmdstanr::install_cmdstan()
```

## Preparing your data

Your data frame must include:

1. **Item response columns** — coded 0, 1, ..., K-1 (0-indexed). Column names
   must follow a recognized lexicon (equate, ne25, credi, gsed, etc.). The
   package auto-detects which lexicon your data uses.
2. **`years_old`** — age in years (e.g., 2.5 for 2 years 6 months).
3. **ID columns** — one or more columns that uniquely identify each observation.

```{r}
library(KidsightsPublic)

# Example: data with equate-lexicon column names
head(my_data)
#>   child_id site years_old AA4 AA5 AA6 AA7 ...
#>   1001     ne25  1.2       0   0  NA   0  ...
#>   1002     ne25  3.5       1   1   1   1  ...
```

## Scoring Kidsights (overall developmental score)

```{r}
results <- score_kidsights(my_data, id_cols = c("child_id", "site"))
head(results)
#>   child_id site  theta
#>   1001     ne25  -2.31
#>   1002     ne25   3.45
```

The `theta` column contains the MAP estimate of the latent trait. Higher values
indicate more advanced development. Observations with fewer than 5 non-missing
item responses receive `NA`.

## Scoring psychosocial domains

```{r}
ps_results <- score_psychosocial(my_data, id_cols = c("child_id", "site"))
head(ps_results)
#>   child_id site theta_gen theta_eat theta_ext theta_int theta_sle theta_soc
#>   1001     ne25  0.12     -0.45      0.33     -0.21      0.55     -0.10
#>   1002     ne25  1.05      0.22      0.87      0.15      0.33      0.68
```

The psychosocial model produces six scores:

| Column | Domain |
|--------|--------|
| `theta_gen` | General psychosocial factor |
| `theta_eat` | Eating problems |
| `theta_ext` | Externalizing behavior |
| `theta_int` | Internalizing behavior |
| `theta_sle` | Sleep problems |
| `theta_soc` | Social competency |

## Lexicon auto-detection

The package automatically detects which lexicon your column names follow. You
can check this manually:

```{r}
lex <- detect_lexicon(my_data)
lex$lexicon
#> [1] "ne25"
lex$n_matched
#> [1] 185
```

Supported lexicons: equate, kidsight, ne25, ne22, ne20, credi, gsed, mn26,
cahmi21, cahmi22, ecdi.

## Notes

- **Age**: must be in raw years. The model computes `ln(age + 1)` internally.
- **Missing data**: handled via full-information maximum likelihood (FIML).
  Missing item responses are skipped in the likelihood. Observations with
  fewer than 5 responses (configurable via `min_responses`) are not scored.
- **Out-of-range responses**: values outside `[0, K-1]` for an item are set to
  `NA` with a warning.
- **Speed**: optimization typically takes 2-15 seconds depending on sample size
  and model.
